---
title: "Mapping Reclamation: From Erasure to Healing"
goal: "Analyze the spatial relationship between former Indian Residential School (IRS) locations and Indigenous-led infrastructure projects to identify signs of reclamation and healing."
authors: "Jennifer Estigene (235264) & Iana Lanceta (235531)"
output: 
  html_document:
    css: Text Style 1.css
date: "2025-05-14"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Setup
## Load Required Packages
```{r}
# Load necessary libraries 
library(sf)          # For handling spatial vector data
library(tidyverse)   # For data manipulation and plotting
library(tmap)        # For thematic mapping
library(janitor)     # For clean column names
library(here)        # Helps manage file paths
library(dbscan)      # For hotspot clustering
library(leaflet)    # For creating interactive web maps
library(mapview)    # For exporting interactive web maps as images
library(spatstat)   # For conducting a cross-K function
library(ggplot2)    # For generating a cross-K function flow map
library(scales)     # For graphing visualizations
```

## Load Data
```{r}
# Set data folder path
data_folder <- "C:/Users/ianal/OneDrive/Documents/Academic/Hertie MPP/2025 Spring Semester/E14_Geospatial Analysis for Data Science/Data Analysis/Data Files"

# Load former residential school locations
irs <- st_read(file.path(data_folder, "IRS_Locations.shp"))

# Load Indigenous community infrastructure points
infrastructure <- st_read(file.path(data_folder, "Indigenous_community_infrastructure.shp"))

# Load Canadian map
canada_map <- st_read(file.path(data_folder, "lpr_000b21a_e.shp"))
```

## Data Cleaning
```{r}
# Check structure of the datasets
glimpse(irs)
glimpse(infrastructure)

# Check CRS (coordinate reference system)
st_crs(irs)
st_crs(infrastructure)

# Reproject both datasets to Canada Lambert Conformal Conic to accurately represent local shapes and directions across much of the country
canada_crs <- 3347
canada_map_3347 <- st_transform(canada_map, crs = 3347)

irs_proj <- st_transform(irs, crs = canada_crs)
infrastructure_proj <- st_transform(infrastructure, crs = canada_crs)

# Check to confirm
st_crs(irs_proj)
st_crs(infrastructure_proj)
```

## Create Buffer Zones
```{r}
# Create buffers at various distances (in meters)
irs_1km <- st_buffer(irs_proj, dist = 1000) %>% mutate(buffer_km = 1)
irs_5km <- st_buffer(irs_proj, dist = 5000) %>% mutate(buffer_km = 5)
irs_10km <- st_buffer(irs_proj, dist = 10000) %>% mutate(buffer_km = 10)
irs_20km <- st_buffer(irs_proj, dist = 20000) %>% mutate(buffer_km = 20)
irs_50km <- st_buffer(irs_proj, dist = 50000) %>% mutate(buffer_km = 50)

# Combine all into a single sf object
irs_buffers <- bind_rows(irs_1km, irs_5km, irs_10km, irs_20km, irs_50km)

```


# Section 1: Primary Analysis
The purpose of this analysis is to explore the geospatial relationship
between former Canadian Indigenous residential school sites (IRS) and
current Indigenous-led infrastructure projects (IIPs). In particular,
point pattern comparisons and Cross-K functions are conducted in order
to determine if areas surrounding former IRS sites are experiencing
notable investment in Indigenous-led infrastructure projects. Such
actions reflect broader efforts at cultural, social, and spatial
reclamation that have been discussed in topic literature.


## High-Level Overview: Projects per Buffer Zone
The below table summarizes the number of IIPs within each defined
"nearby" buffer of 1, 5, 10, 20, and 50km, as well as the number of IIPs
outside this threshold.
```{r}
# Spatial join: match infrastructure points to buffers
infrastructure_joined <- st_join(infrastructure_proj, irs_buffers, join = st_within)

# Count how many infrastructure projects fall within each buffer
infrastructure_counts <- infrastructure_joined %>%
  st_drop_geometry() %>%
  group_by(buffer_km) %>%
  summarise(projects_within = n())

infrastructure_counts

```

While the results reveal that the number of IIPs increases with
distance,the cumulative number of projects within 50 km of former IRS
sites is greater than those beyond.


## Buffer Zone Composition by Project Category
Since "infrastructure" projects can include both practical municipal
installments (such as water and waste management) as well as
socio-cultural community development initiatives (such as those related
to recreation, health, and education), it is meaningful to investigate
what types of projects dominate within different distance bands.
```{r}
# Summarize: Count by category and buffer
infrastructure_by_category <- infrastructure_joined %>%
  st_drop_geometry() %>%
  group_by(buffer_km, TYPE_EN) %>%
  summarise(projects = n(), .groups = "drop") %>%
  arrange(buffer_km, desc(projects))

infrastructure_by_category

```

The insights within this table suggest that spatial proximity may more
closely reflect symbolic reclamation compared to practical
infrastructure delivery.


## Figure 1: Overlaying Buffer Zone Counts with Project Type Volumes
```{r}
Figure_1 <- ggplot(infrastructure_by_category, aes(x = factor(buffer_km), y = projects, fill = TYPE_EN)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Buffer Distance (km)",
    y = "Number of Indigenous-Led Infrastructure Projects",
    title = "Figure 1- IIP by Buffer & Project Type",
    fill = "Project Type",
  ) +
  scale_x_discrete(labels = c("1", "5", "10", "20", "50", "> 50km")) +
  scale_y_continuous(labels = comma) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    axis.title.x = element_text(size = 10, margin = margin(t = 5)), 
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.line = element_line(color = "black", linewidth = 0.5),   
    axis.ticks = element_line(color = "black", linewidth = 0.5),
    plot.title = element_text(hjust = 0.5, vjust = 2, size = 11), 
    legend.title = element_text(size = 10) 
  )

Figure_1

# Save the plot as a PNG file
ggsave("Figure 1- IIP by Buffer & Project Type.png", plot = Figure_1, width = 10, height = 6, units = "in", dpi = 300)
```

This stacked bar graph demonstrates the findings noted above: (1) IIP
volume increases with buffer zone level, (2) total IIPs within 50 km of
IRS sites is greater than those beyond this threshold, and (3) there is
a considerable amount of socio-cultural projects (brown to teal colours)
that hint at symbolic forms of spatial reclamation.


## Figure 2: Map of High-Reclamation IRS Sites
For future policy discussions and program development, it is useful to
know where the "most reclaimed" IRS site are located across Canada.
```{r}
# Count projects within 10 km of each IRS site
irs_10km <- st_buffer(irs_proj, dist = 10000) %>%
  mutate(irs_id = row_number())

projects_per_site <- st_join(infrastructure_proj, irs_10km, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(irs_id) %>%
  summarise(n_projects = n(), .groups = "drop")

# Select top 5 IRS sites by number of nearby projects
top_reclaimed_ids <- projects_per_site %>%
  arrange(desc(n_projects)) %>%
  slice_head(n = 5) %>%
  pull(irs_id)

# Highlight these on the map
top_irs_sites <- irs_proj %>%
  mutate(irs_id = row_number()) %>%
  filter(irs_id %in% top_reclaimed_ids)

# Map all IRS sites in gray, top 5 in red
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(data = st_transform(irs_proj, 4326),  # all sites
                   radius = 3, color = "gray", stroke = FALSE, fillOpacity = 0.6) %>%
  addCircleMarkers(data = st_transform(top_irs_sites, 4326),  # top 5
                   radius = 6, color = "red", stroke = TRUE, fillOpacity = 0.9,
                   label = ~paste("Top IRS site", irs_id)) %>%
  setView(lng = -95, lat = 55, zoom = 3)

leaflet_map <- leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(data = st_transform(irs_proj, 4326),  # all sites
                   radius = 3, color = "gray", stroke = FALSE, fillOpacity = 0.6) %>%
  addCircleMarkers(data = st_transform(top_irs_sites, 4326),  # top 5
                   radius = 6, color = "red", stroke = TRUE, fillOpacity = 0.9,
                   label = ~paste("Top IRS site", irs_id)) %>%
  setView(lng = -95, lat = 55, zoom = 3)


# Save the map into a static image 
Figure_2 <- ggplot() +
  geom_sf(data = canada_map_3347, fill = "lightgray", color = "black") +
  geom_sf(data = st_transform(irs_proj, crs = st_crs(canada_map_3347)),
          aes(color = "All IRS Sites"), size = 1) +
  geom_sf(data = st_transform(top_irs_sites, crs = st_crs(canada_map_3347)),
          aes(color = "Top 5 IRS Sites"), size = 3) +
  scale_color_manual(values = c("All IRS Sites" = "blue", "Top 5 IRS Sites" = "red"),
                     name = "IRS Site Category") + 
  coord_sf(crs = 3347) +
  labs(title = "Figure 2- Top IRS Sites by Nearby Infrastructure Projects") + 
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, vjust = 2, size = 11),
    legend.position = "bottom") 

Figure_2

# Save the plot as a PNG file
ggsave("Figure 2- Map of High Reclamation IRS Sites.png", plot = Figure_2, width = 10, height = 6, units = "in", dpi = 300)

```

Rather than concentrated in a specific region, the top 5 most reclaimed
IRS sites are spread across the country. This knowledge can serve as the
starting point for further research to see how effectively other truth
and reconciliation efforts are implemented in these areas.


## Cross-K Analysis
Conducting a cross-k function will help identify any statistically
significant correlations between point patterns by checking if point
patterns cluster at the same locations.


### Figure 3: Spatial Statistics
The plot shows the estimated cross-K function,
$\hat{K}_{irs,infrastructure}(r)$ (representing the main result: cross-K
function between IRS and infrastructure), as the solid black line
(labeled $\hat{K}_{irs,infrastructure}^{iso}(r)$ which stands for
isotropic correction, a type of edge correction that tries to account
for this bias in a way that assumes the pattern is the same in all
directions. The x-axis represents the distance $r$, and the y-axis
represents the value of the cross-K function at that distance.

The other lines on the plot represent different edge correction methods
($\hat{K}_{irs,infrastructure}^{trans}(r)$ for translation correction
and $\hat{K}_{irs,infrastructure}^{bord}(r)$ for border correction). As
can be seen, these should lines are reasonably close to each other-
which is ideal. The blue dashed line
($\hat{K}_{irs,infrastructure}^{pois}(r)$) represents the theoretical
value of the cross-K function under the null hypothesis of complete
spatial randomness (CSR) and independence between the two point
patterns.
```{r}
# Extract bounding boxes from the sf objects
irs_bbox <- st_bbox(irs)
infrastructure_bbox <- st_bbox(infrastructure)

# Determine the overall extent by finding the minimum and maximum of all coordinates
xmin <- min(irs_bbox["xmin"], infrastructure_bbox["xmin"])
xmax <- max(irs_bbox["xmax"], infrastructure_bbox["xmax"])
ymin <- min(irs_bbox["ymin"], infrastructure_bbox["ymin"])
ymax <- max(irs_bbox["ymax"], infrastructure_bbox["ymax"])

# Define the observation window
window <- owin(c(xmin, xmax), c(ymin, ymax))

# Create ppp objects using the extracted coordinates
irs_coords <- st_coordinates(irs)
infrastructure_coords <- st_coordinates(infrastructure)

irs_ppp <- ppp(x = irs_coords[, 1], y = irs_coords[, 2], window = window)
infrastructure_ppp <- ppp(x = infrastructure_coords[, 1], y = infrastructure_coords[, 2], window = window)

# Combine coordinates
all_x <- c(irs_coords[, 1], infrastructure_coords[, 1])
all_y <- c(irs_coords[, 2], infrastructure_coords[, 2])

# Create a vector of marks indicating the type of each point
all_types <- c(rep("irs", nrow(irs_coords)), rep("infrastructure", nrow(infrastructure_coords)))

# Create a multitype ppp object
all_points_marked <- ppp(x = all_x, y = all_y, window = window, marks = factor(all_types))

# Define a sequence of distances (r)
max_distance <- diff(range(window$x)) / 5
r_values <- seq(0, max_distance, length.out = 100)

# Calculate the cross-K function
cross_K <- Kcross(all_points_marked, i = "irs", j = "infrastructure", r = r_values)

plot(cross_K,
     main = "Figure 3- Spatial Statistics: Former IRS and IIPs",
     cex.main = 0.8,
     cex.lab = 0.7,
     cex.axis = 0.6,
     xlab = "r (distance in km)",
     ylab = expression(K[irs, infrastructure](r)),
     legend = FALSE,
     mgp = c(1.5, 0.5, 0)
)

# Adding and formatting legend
legend("topleft",
       legend = c(expression(hat(K)[irs, infrastructure]^{iso}(r)),
                  expression(hat(K)[irs, infrastructure]^{trans}(r)),
                  expression(hat(K)[irs, infrastructure]^{bord}(r)),
                  expression(K[irs, infrastructure]^{pois}(r))),
       lty = c(1, 2, 3, 4),
       col = c("black", "red", "forestgreen", "blue"),
       cex = 0.6
)

```

*This plot indicates that former residential schools and Indigenous
community infrastructure tend to be located closer to each other more
often than you would expect if their locations were completely random
and independent of each other*. This clustering effect seems to be more
pronounced at distances up to around 3km from a former IRS and current
infrastructure project. Beyond the 5km mark, the spatial relationship
appears to weaken.

The below code is simply to export the above graph.
```{r}
# Printing as PNG
dev.off()

# Print a confirmation message
print("Cross-K function plot saved as cross_k_function.png")
```


### Figure 4: Geographically Weighted Flow Cross-K Function
A geographically weighted flow cross-k function in order to see the
previous findings visualized as a map.
```{r}
# Create map using ggplot2
ggplot() +
  geom_sf(data = irs, aes(color = "Former IRS"), size = 1.5, shape = 16) +
  geom_sf(data = infrastructure, aes(color = "Indigenous Infrastructure"), size = 1.5, shape = 17) +
  scale_color_manual(name = "Point Type",  # Set the legend title
                     values = c("Former IRS" = "red", "Indigenous Infrastructure" = "blue")) +
  labs(title = "Figure 4- Geographically Weighted Flow Cross-K Map") +
  theme_minimal() +
  theme(legend.position = "bottom")

leaflet() %>%
  addTiles() %>% # Add default OpenStreetMap tiles
  addCircleMarkers(data = irs,
                   color = "red",
                   radius = 5,
                   fillOpacity = 0.8,
                   label = "Former IRS") %>%
  addCircleMarkers(data = infrastructure,
                   color = "blue",
                   radius = 5,
                   fillOpacity = 0.8,
                   label = "Indigenous Infrastructure") %>%
  addLegend(position = "bottomright",
            colors = c("red", "blue"),
            labels = c("Former IRS", "Indigenous Infrastructure"))

irs_transformed <- st_transform(irs, crs = st_crs(canada_map_3347))
infrastructure_transformed <- st_transform(infrastructure, crs = st_crs(canada_map_3347))

# Save map into a static image
Figure_4 <- ggplot() +
  geom_sf(data = canada_map_3347, fill = "lightgray", color = "black") +
  geom_sf(data = irs_transformed, aes(color = "Former IRS"), size = 1.5, shape = 16) +
  geom_sf(data = infrastructure_transformed, aes(color = "Indigenous Infrastructure"), size = 1.5, shape = 17) +
  scale_color_manual(name = "Point Type",  # Set the legend title
                     values = c("Former IRS" = "red", "Indigenous Infrastructure" = "blue")) +
  labs(title = "Figure 4- Geographically Weighted Flow Cross-K Map") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Save the ggplot2 map as a PNG file
ggsave("Figure 4- Geographically Weighted Flow Cross-K Map.png", plot = Figure_4, width = 10, height = 8, units = "in", dpi = 300)

# Confirmation message
print("Figure 4- Geographically Weighted Flow Cross-K Map.png")

```

As described above, there are clear clusters of IIPs atop where former
IRS sites used to be. In line with topic literature (McBain, 2021), this
analysis finds that former IRS sites are experiencing notable investment
in nearby IIPs, which reflects broader efforts at cultural, social, and
spatial land reclamation. These transformations are vital to healing as
they encourage reconnection with erased cultural landscapes [and
challenge] current Eurocentric conservation practices” (McBain, 2021).


# Section 2: Data Sources/References

Canadian Map – Provinces & Territories Dataset (DOI:
<https://www12.statcan.gc.ca/census-recensement/alternative_alternatif.cfm>)

Former IRS site data from the Borealis Data Repository (DOI:
10.5683/SP2/FJG5TG)

Indigenous Services Canada’s IIP Dataset (DOI: geo.sac-isc.gc.ca)

McBain, C. (2021). De-Colonial Intersections of Conservation and
Healing: The Indian Residential School System (Doctoral dissertation,
Carleton University).


# Section 3: Appendix- Further Analysis
The scope of this assignment and research question limits the number of
figures included in the primary analysis. However, given the importance
of reconciliation and reclamation efforts in Canada today, the below
analyses contribute to research in the space to support healing and
data-driven policymaking and program development.


## Buffer Breakdown by Province/Territory
Question: Are some provinces more active in reclaiming land near former
IRS sites?
```{r}
# Group and count by province and buffer distance
infrastructure_by_province <- infrastructure_joined %>%
  st_drop_geometry() %>%
  group_by(PROV_EN, buffer_km) %>%
  summarise(projects = n(), .groups = "drop") %>%
  arrange(PROV_EN, buffer_km)

infrastructure_by_province
```

This table breaks down infrastructure projects by buffer distance and
province, revealing spatial variation across regions. Some provinces,
such as British Columbia and Alberta, show strong project clustering
near IRS sites, while others show more dispersed patterns - reflecting
regional differences in historical settlement, school density, and
funding allocation.


## Former Schools with Highest Concentration Of Surrounding Projects
```{r}
# Create count of IRS sites per province (using PROV_EN)
irs_by_province <- irs_proj %>%
  st_drop_geometry() %>%
  rename(PROV_EN = Province) %>%   # ensure naming matches
  group_by(PROV_EN) %>%
  summarise(n_schools = n(), .groups = "drop")

# Create lookup table
prov_lookup <- tibble(
  PROV_EN = c("AB", "BC", "MB", "NB", "NL", "NS", "NT", "NU", "ON", "PE", "QC", "SK", "YT"),
  province_full = c(
    "Alberta", "British Columbia", "Manitoba", "New Brunswick", "Newfoundland and Labrador",
    "Nova Scotia", "Northwest Territories", "Nunavut", "Ontario", "Prince Edward Island",
    "Quebec", "Saskatchewan", "Yukon"
  )
)

# Add province_full to both datasets
infrastructure_by_province_named <- infrastructure_by_province %>%
  left_join(prov_lookup, by = "PROV_EN")

irs_by_province_named <- irs_by_province %>%
  left_join(prov_lookup, by = "PROV_EN")

# Summarize IRS sites per province (1 row per province_full)
irs_by_province_summarized <- irs_by_province_named %>%
  group_by(province_full) %>%
  summarise(n_schools = sum(n_schools), .groups = "drop")

# Join and calculate projects per school
province_normalized <- infrastructure_by_province_named %>%
  left_join(irs_by_province_summarized, by = "province_full") %>%
  mutate(projects_per_school = round(projects / n_schools, 2))

# View result
province_normalized
```

This table normalizes infrastructure project counts by the number of IRS
sites per province, highlighting the relative intensity of project
activity. Provinces with higher "projects per school" values suggest
more spatially concentrated reclamation efforts. This metric reveals
where Indigenous communities may be focusing more actively on reclaiming
space near former school locations.

```{r}
# Unique values in each dataset
unique(infrastructure_by_province$PROV_EN)
unique(irs_by_province$PROV_EN)
```


## IRS Site-Level Reclamation Density
Question: Which former schools have the highest concentration of
surrounding projects?
```{r}
# Create 10km buffer around each IRS site with unique ID
irs_10km <- st_buffer(irs_proj, dist = 10000) %>%
  mutate(irs_id = row_number())

# Join projects to IRS sites using the 10km buffer
projects_per_site <- st_join(infrastructure_proj, irs_10km, join = st_within)

# Count projects per IRS site
projects_per_site_summary <- projects_per_site %>%
  st_drop_geometry() %>%
  group_by(irs_id) %>%
  summarise(projects_nearby = n(), .groups = "drop") %>%
  arrange(desc(projects_nearby))

projects_per_site_summary

# Join back with original IRS data to get names
projects_per_site_summary_named <- projects_per_site_summary %>%
  left_join(
    irs_proj %>% mutate(irs_id = row_number()) %>% st_drop_geometry() %>%
      select(irs_id, NRCNamePri),
    by = "irs_id"
  ) %>%
  select(NRCNamePri, projects_nearby) %>%
  arrange(desc(projects_nearby))

projects_per_site_summary_named
```

This summary identifies former IRS sites with the most and fewest nearby
infrastructure projects (within 10km). The variation between sites
highlights uneven spatial patterns of reclamation. Sites with no or few
nearby projects may represent under-recognized areas of potential
cultural revitalization or indicate differing local priorities.


## Distance-to-Nearest IRS Site
Question: How close is each infrastructure project to the nearest former
IRS site?
```{r}
# Calculate pairwise distances (in meters)
distance_matrix <- st_distance(infrastructure_proj, irs_proj)

# Get the minimum distance to any IRS site for each project
infrastructure_proj$min_distance_m <- apply(distance_matrix, 1, min)

# Convert meters to kilometers
infrastructure_proj$min_distance_km <- round(infrastructure_proj$min_distance_m / 1000, 2)

# Summarize distance statistics
infrastructure_proj %>%
  st_drop_geometry() %>%
  summarise(
    mean_distance_km = mean(min_distance_km),
    median_distance_km = median(min_distance_km),
    min_distance_km = min(min_distance_km),
    max_distance_km = max(min_distance_km)
  )

infrastructure_proj
```

This records how far each project is from the nearest IRS site. It
allows us to assess overall proximity and evaluate whether projects tend
to cluster near former schools. The majority of projects appear within
50km, reinforcing the spatial relationship between reclamation activity
and former IRS locations.


## Density Ratios: Projects Per Square KM
Question: Are projects more spatially dense near IRS sites, or further
away?
```{r}
# Calculate total area per buffer_km (in km²)
buffer_areas <- irs_buffers %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) %>%
  st_drop_geometry() %>%
  group_by(buffer_km) %>%
  summarise(total_area_km2 = sum(area_km2), .groups = "drop")

# Reuse the infra_counts data (number of projects per buffer_km)
# Make sure infra_counts exists from previous steps

# Join counts with areas and compute density
density_summary <- left_join(infrastructure_counts, buffer_areas, by = "buffer_km") %>%
  mutate(
    projects_per_km2 = round(projects_within / total_area_km2, 4)
  ) %>%
  arrange(buffer_km)

density_summary
```

This table calculates project density (projects per square kilometer)
for each buffer zone. It shows a steep decline in density as distance
increases from the IRS sites - indicating a strong spatial concentration
of projects in areas closest to former schools. This supports the
hypothesis that land reclamation efforts are geographically targeted
toward these historical sites.


## Buffer Ring Analysis
Question: How does the number of infrastructure projects change across
distance bands from former IRS sites?
```{r}
# Ensure counts are sorted
infrastructure_ring <- infrastructure_counts %>%
  arrange(buffer_km) %>%
  mutate(
    ring_label = case_when(
      buffer_km == 1 ~ "0–1 km",
      buffer_km == 5 ~ "1–5 km",
      buffer_km == 10 ~ "5–10 km",
      buffer_km == 20 ~ "10–20 km",
      buffer_km == 50 ~ "20–50 km",
      TRUE ~ NA_character_
    ),
    ring_projects = projects_within - lag(projects_within, default = 0)
  )

infrastructure_ring
```

This analysis separates project counts into distinct distance rings
(e.g., 0-1km, 1-5km, 5-10km, etc.), rather than cumulative buffers. The
results show a sharp drop-off in the number of projects as distance
increases, with the highest concentration occurring closest to former
IRS sites. This supports the hypothesis that many infrastructure efforts
are spatially clustered around these historical locations, potentially
reflecting deliberate reclamation.


## Top 5 Project Types per Distance Band
Question: Which types of infrastructure projects are most commonly
located near former IRS sites, and how does this vary by distance?
```{r}
top_types_per_buffer <- infrastructure_by_category %>%
  group_by(buffer_km) %>%
  slice_max(order_by = projects, n = 5) %>%
  arrange(buffer_km, desc(projects))

top_types_per_buffer
```

This breakdown identifies the top five project types within each buffer
distance from former IRS sites. It reveals how certain types of
infrastructure - such as housing or education facilities - are more
frequently placed in close proximity to these historical locations. This
pattern suggests that spatial placement may reflect symbolic intent,
service delivery priorities, or funding strategies tied to proximity.


## Compare "Ongoing" vs "Completed" Project Proximity
Question: Do ongoing infrastructure projects tend to be located closer
to former IRS sites than completed ones?
```{r}
infrastructure_proj %>%
  st_drop_geometry() %>%
  filter(!is.na(min_distance_km)) %>%
  group_by(STATUS_EN) %>%
  summarise(
    avg_distance = mean(min_distance_km),
    median_distance = median(min_distance_km),
    count = n()
  )

infrastructure_proj
```

This comparison of average and median distances shows that ongoing
projects are, on average, closer to former IRS sites than completed
ones. While the dataset lacks full timestamps, project status offers a
temporal proxy. The results suggest that more recent infrastructure
efforts are increasingly concentrated near IRS sites- indicating a
growing spatial focus on land reclamation and healing in these
historically significant areas.


## Under-Reclaimed IRS Sites (Few or No Projects Nearby)
Question: Which former IRS sites have few or no infrastructure projects
within 10km?
```{r}
# Buffer around IRS sites
irs_10km <- st_buffer(irs_proj, dist = 10000) %>%
  mutate(irs_id = row_number())

# Join and count projects per IRS site
low_reclaimed <- st_join(infrastructure_proj, irs_10km, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(irs_id) %>%
  summarise(n_projects = n()) %>%
  right_join(irs_proj %>% mutate(irs_id = row_number()), by = "irs_id") %>%
  replace_na(list(n_projects = 0)) %>%
  filter(n_projects <= 2) %>%
  select(NRCNamePri, n_projects) %>%
  arrange(n_projects)

low_reclaimed
```

This table identifies IRS sites with few or no nearby infrastructure
projects within 10km. These locations may represent under-recognized or
under-served areas.


## Spatial Proximity Analysis
Question: How can infrastructure projects be categorized based on their
spatial proximity to former IRS sites, and what insights emerge from
comparing these categories?
```{r}
# Add distance category labels to infrastructure projects
infrastructure_proj <- infrastructure_proj %>%
  mutate(distance_band = case_when(
    min_distance_km <= 5 ~ "Close (≤5km)",
    min_distance_km <= 20 ~ "Mid (5–20km)",
    TRUE ~ "Far (>20km)"
  ))

# View breakdown by band
infrastructure_proj %>%
  st_drop_geometry() %>%
  count(distance_band) %>%
  mutate(percent = round(100 * n / sum(n), 1))
```

Projects were grouped into three distance bands to simplify proximity
analysis. This categorization enables clearer comparisons across project
types, provinces, and statuses. The majority of projects fall into the
"Far" category, but a notable concentration is within the "Close" range
supports the presence of targeted reclamation near former school sites.


## Project Type Distribution by Proximity Category

Question: Which types of infrastructure projects are most common within
close, mid, and far distances from former IRS sites?

```{r}
# Project types by proximity band
proj_type_by_distance <- infrastructure_proj %>%
  st_drop_geometry() %>%
  group_by(distance_band, TYPE_EN) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(distance_band, desc(n))

proj_type_by_distance
```

This analysis reveals that certain types of infrastructure projects are
disproportionately represented in areas close to former Indian
Residential School (IRS) sites. For instance, cultural, education, and
health-related projects are most common in the “Close (≤5 km)” band,
suggesting that these forms of development may be spatially targeted to
reclaim and revitalize historically significant sites. In contrast,
broader categories such as general housing or transportation
infrastructure are more evenly distributed or tend to increase with
distance, possibly reflecting regional infrastructure needs beyond
symbolic reclamation.

## Project Types Are Closest to IRS Sites
Question: Which types of infrastructure projects are spatially closest
(on average) to former IRS sites?
```{r}
# Analyze average proximity to IRS sites by project type
proj_type_proximity <- infrastructure_proj %>%
  st_drop_geometry() %>%
  group_by(TYPE_EN) %>%
  summarise(
    avg_distance_km = mean(min_distance_km, na.rm = TRUE),
    n_projects = n()
  ) %>%
  arrange(avg_distance_km)

# View results: which project types are closest on average
proj_type_proximity
```

When sorted by average distance to the nearest former IRS site, the
analysis shows that health, and service based infrastructure projects
are among the closest. These types of projects have the smallest mean
distance values, indicating a strong spatial and symbolic connection to
the legacy of the residential school system. The proximity of these
project types supports the view that certain infrastructure categories
are intentionally placed near former school locations as a form of
land-based healing or remembrance.
